Best Practices in Latent Class Analysis using Free Open Source Software

### 2023-02-16 manuscript.tex

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{latent class analysis, mixture models, best practices, free open source software, tidySEM\newline\indent Word count: 9001}
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Best Practices in Latent Class Analysis using Free Open Source Software},
  pdfauthor={Caspar J. van Lissa1,2, Mauricio Garnier-Villarreal3, \& Daniel Anadria1},
  pdflang={en-EN},
  pdfkeywords={latent class analysis, mixture models, best practices, free open source software, tidySEM},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

LCA can also be used as a measurement model, similar to confirmatory factor analysis.
This is useful when class membership is measured imperfectly by several indicators:
LCA can partial out measurement error and restore most likely class membership.
For example, one study found discrepancies between self-reported employment status and official register data (Pavlopoulos \& Vermunt, 2015).
While it might be tempting to assume that register data is free of error, this assumption was shown to be false.
A mixture model was used to estimate employment status, correcting for measurement error, from both self-report and registry data.
The study investigated random and systematic measurement error in the indicatorss stemming from the two registers,
which revealed strengths and weaknesses of both types of assessment.

@@

The use of LCA for dimension reduction is common in longitudinal research,
where one developmental process is summarized using latent class growth analysis (LCGA),
and the resulting categories are then used as a moderator of another developmental process in a multi-group model.
For example, one study conducted a LCGA of adolescent empathy development on two dimensions of empathy identifying: high, average, and low empathy classes (Van Lissa et al., 2015).
The membership to these three groups was then used as a moderator in a latent growth analysis of adolescent- and parent-reported conflict.
The results showed that the high-empathy group evidenced greater adolescent-parent agreement about the incidence of conflict than other groups,
and the low empathy group had more conflict with parents than other groups.
Note that as both of these effects were non-linear,
modeling them in a single-group framework would introduce unnecessary bias.
LCGA greatly simplifies model specification and interpretation,
albeit at a cost of some loss of information, and potentially generalizability.
Appendix C is a tutorial on LCGA,
based on ongoing research on trajectories of depressive symptoms in military service members after deployment.

The need for dimension reduction can be especially pertinent when using ordinal indicators.
For example, one study measured
fifteen dichotomous health indicators among injured military personnel (MacGregor et al., 2021).
Combinatorics informs us that fifteen
dichotomous items have \(2^{15}\) or \(32,768\) unique symptom combinations.
However, LCA was able to reduce response patterns to five clinically meaningful latent classes supported by the data.

@@

\hypertarget{violations-of-normality}{%
\subsubsection{Violations of normality}\label{violations-of-normality}}

Finally, LCA can be used to deal with data which violate certain assumptions.
For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) {[}Baughman (2007); magnus\_multidimensional\_2021{]}.
These models estimate one class to explain the ``true zeros'',
and another class with a different distribution to explain all other values, which can include zero.
Note that this LCA has different models for both classes,
and thus falls beyond the scope of this paper.
For instance, the aforementioned study on marine plastic noted that many prior studies had treated the distribution of particles as (multivariate) normal, despite clear deviations from normality (Alkema et al., 2022).
A mixture of two normal distributions with distinct variances described the observed distribution better.

@@

The best practices in estimation, as outlined in Table
\ref{tab:checkest}, are rooted in existing recommendations for best
practices for estimating specific subtypes of LCA,
including latent
class growth analysis (Van De Schoot, Sijbrandij, Winter, Depaoli, \& Vermunt, 2017) and latent
class analysis with ordinal indicators (e.g., Nylund-Gibson \& Choi, 2018).
These were generalized to be relevant to all types of LCA, and updated to current best practices, as explained below.

@@

\begin{tabular}[t]{l|l}
\hline
\# & Item\\
\hline
1. & Examining Observed Data\\
\hline
2. & Handling Missing Data\\
\hline
3. & Alternative Model Specifications\\
\hline
4. & Software\\
\hline
5. & Algorithm\\
\hline
6. & Class Enumeration\\
\hline
7. & Model Fit Indices\\
\hline
8. & Classification Diagnostics\\
\hline
9. & Interpreting Class Solutions\\
\hline
10. & Label switching\\
\hline
\end{tabular}

@@

Examining observed data is essential for any analysis as it may reveal
patterns and violations of assumptions that had not been considered
prior to data collection.
The function \texttt{descriptives(data)} provides extensive descriptive statistics,
including the number of unique values,
variance of continuous and categorical variables,
missingness, skew and kurtosis (peakedness of the distribution).
However, sample-level descriptive statistics are of limited value when the goal of a study is to identify subsamples using LCA.
Plots convey additional information; specifically, density plots for continuous variables, and bar charts
for categorical ones.
Density plots can also aid in the choice of the number of classes, as explained in the section on visualization.

Particular points of attention include measurement level of the indicators.
Recall that \texttt{tidySEM} can estimate mixture models with either numeric (type \texttt{numeric} and \texttt{integer}) or \texttt{ordered} indicators.
The column \texttt{type} should thus only contain these types.
Also check that the number of \texttt{unique} values match the intended measurement level.
Numeric variables with few unique values are typically better modeled as ordinal.
Our experience consulting on LCA methods
and moderating the \texttt{tidyLPA} Google group suggest that the
misapplication of mixture models to ordinal (e.g., Likert-type)
indicators is the most common source of user error. Whereas it has been
argued that some parametric methods are robust when scales with 7+
response categories are treated as continuous {[}e.g., Norman (2010); rhemtulla\_when\_2012{]},
it is very unlikely
that an ordinal variable can be modeled as a \emph{mixture} of multiple
normal distributions.
To illustrate the problem, consider what happens when estimating a 5-class solution on a single 5-point Likert scale indicator.
Each class-specific mean would describe a single response
category,
and a class-specific variance component would be empirically underidentified.
In sum, variables with few unique values might not be suitable for Gaussian mixture modeling;
LCA with ordinal indicators is more appropriate.

\hypertarget{data-preprocessing}{%
\subsubsection{Data preprocessing}\label{data-preprocessing}}

LCA can accommodate variables of different measurement levels: continuous, binary, nominal, and ordinal.
Numeric variables should have an interval measurement level,
such that it is sensible to estimate means and (co)variances.
For the remaining three measurement levels,
some preprocessing is required.
The exact data types accepted by \texttt{tidySEM} at the time of writing are \texttt{numeric} and \texttt{ordered}.
To correctly specify ordered variables
such as Likert scales,
use the function \texttt{OpenMx::mxFactor()}.
Binary variables,
such as presence or absence of some symptom, are also treated as ordinal, with one category coded as zero and the other as one.
Nominal variables are converted to binary indicators of group membership via dummy coding, for example, using \texttt{mx\_dummies()}.

@@

A third consideration in preprocessing relates to the univariate distributions of indicators.
For example, the \texttt{plas\_depression} data show extreme right skew in depressive symptoms:
most respondents report few symptoms,
and few respondents report many symptoms.
Similarly, the \texttt{alkema\_microplastics} data are extremely right-skewed: Very small plastic fragments abound in the oceans, and large pieces are rare.
These distributional characteristics must be taken into account either in preprocessing, or model specification.
In preprocessing, the data can be transformed to reduce skew.
After reducing skew, a simpler LCA model may fit the data well.
Alternatively,
one can specify a model that accommodates the distributional idiosyncrasies.
For example, allowing free variances across classes might result in one class with a near-zero mean and small variance to account for the fact that most respondents have few depressive symptoms,
and one class with an elevated mean and large variance to account for all respondents with non-zero symptoms.

@@

Previous work has emphasized the importance of examining the pattern of
missing data and reporting how missingness was handled
(Van De Schoot et al., 2017).
While we concur that investigating missingness is due diligence,
it is important to emphasize that missingness is adequately handled by default in \texttt{OpenMx}, which is the computational backend of \texttt{tidySEM}.
Its Full Information Maximum Likelihood (FIML) estimator makes use of all available
information without imputing missing values, by estimating the model based on all
available information for each subject {[}C. Enders and Bandalos (2001); enders\_applied\_2022{]}.
FIML is a best-practice solution for handling missing data;
at least on par with multiple imputation (Lee \& Shi, 2021).
Multiple imputation is less suitable to LCA for two reasons. First,
because LCA methods are often computationally intensive,
and conducting them on several imputed datasets multiplies the load.
Second, because many multiple imputation methods assume normality; an assumption likely to be violated when data come from a heterogeneous population, so the imputation model could bias the results as it does not account for the expected heterogeneity.

@@

A second consideration in model specification is the risk of overfitting (Hastie, Tibshirani, \& Friedman, 2009).
Overfitting occurs when a model has many parameters relative to the number of observations,
causing it to capture idiosyncratic noise in the sample and limiting its generalizability to new samples.
As explained before, the number of parameters in LCA scales with the number of estimated classes.
Consequently, LCA with many classes have a potentially very high number of parameters.
In Bayesian mixture modeling,
it was observed that if the estimated number of classes exceeded the true number, the number of observations assigned to these classes tended towards zero as sample size increased (Rousseau \& Mengersen, 2011).
Although it is not known whether frequentist LCA exhibits the same behavior,
low class counts are still an indicator of potential overfitting.
The section on Class Enumeration addresses ways to guard against overfitting.

@@

As there are no absolute fit indices for LCA,
information criteria are typically used to assess relative fit.
Information criteria are computed from the \(-2LL\),
which is the most basic fit measure, testing the overall hypothesis if the model properly represents the data.
Information criteria add a penalty for model complexity to the \(-2LL\), and thus balance fit and complexity.
The lower
the value of an information criterion, the better the overall fit of the
model.
The original information criterion is the Akaike Information Criterion (AIC),
computed as \(-2LL+2p\), where \(p\) is the number of parameters.
The Bayesian Information Criterion (BIC) multiplies the penalty for complexity with a function of the sample size, \(p \ln(n)\),
such that complex models are penalized more in smaller samples.
Finally,
the so-called \emph{sample size adjusted} BIC (saBIC) implements a different ad-hoc penalty for sample size,
\(\ln(\frac{n + 2}{24})\).
Although simulation studies have showed that the saBIC performed well for class enumeration (Chen, Luo, Palardy, Glaman, \& McEnturff, 2017; Tein, Coxe, \& Cham, 2013),
its penalty is not based on theory.
BIC also performs well and is theoretically substantiated;
as such, it may be the most appropriate information criterion to use for model comparison (Masyn, 2013; Nylund-Gibson \& Choi, 2018).
All three ICs are available in the output of \texttt{table\_fit()}.

Information criteria may occasionally contradict each other, so it is
important to identify a suitable strategy to reconcile them.
One option is to select a specific fit index, such as the BIC,
before seeing the results.
This choice can be preregistered to avoid concerns about cherry picking.
Another option is to always prefer the most parsimonious model that has best fit according to any of the available fit indices. Other approach is to select several competing model based on the ICs, and evaluate the theoretical interpretation.
Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process (Akogul \& Erisoglu, 2016),
which is implemented in \texttt{tidyLPA} (Rosenberg, Beymer, Anderson, Van Lissa, \& Schmidt, 2018).
Finally, one can use a scree plot to determine the inflection point at which additional classes contribute little to further decreases of the ICs (Nylund-Gibson \& Choi, 2018).
The preferred number of classes is at the inflection point where the slope of the curve is clearly leveling off.
Adding this many classes results in substantial decreases of the IC,
and adding further classes results in relatively small decreases.

Given a specific IC,
it is also possible to calculate so-called ``IC-weights'',
which indicate the relative support the data provide for each model in the set (Wagenmakers \& Farrell, 2004).
This is particularly useful for confirmatory LCA,
as it allows one to quantify the support for the theoretical model.
IC-weights are calculated by calling \texttt{ic\_weights()} on the model fit table.
Note, however, that IC-weights will (strongly) favor the model with the lowest IC value;
the scree plot approach above may be more suitable to select the most parsimonious model with adequate fit.

@@

When two models are nested, the significance of the difference in fit between these models can be tested using a Likelihood Ratio test (LR).
The test statistic is computed as the -2 times the log of the ratio of likelihoods,
and its degrees of freedom are equal to the difference in the degrees of freedom of the two compared models.
Since a \(k-1\)-class model is nested in a \(k\)-class model,
it is possible to test the difference in model fit using a LR test.
One challenge is that the LR test statistic is not asymptotically \(\chi^2\) distributed for LCA.
The so-called Lo-Mendell-Rubin adjusted LR test (LMR) overcomes this problem by applying an ad-hoc correction to the LR (Lo, Mendell, \& Rubin, 2001).
This test is implemented in \texttt{tidySEM}, and is part of the default output of \texttt{table\_fit()}, comparing subsequent rows of the table.

Another solution is to derive an empirical sampling distribution for the LR difference.
The so-called bootstrapped Likelihood Ratio Test (BLRT) does so by simulating data from the \(k-1\)-class model \(B\) times (e.g., 1000),
and fitting the \(k\)- and \(k-1\)-class models to each simulated dataset (Nylund et al., 2007).
The p-value of this test is the proportion of \(B\) in which the LR of the models fit to simulated data exceeds the LR observed in the real data.
Simulation studies suggest that the BLRT has greater statistical power than the LRT,
although both performed comparably well at detecting the true number of classes (Nylund et al., 2007; Tein et al., 2013).
A crucial limitation of the BLRT is that it is very computationally expensive.
With this in mind, \texttt{table\_fit()} does not provide the BLRT by default.
Specific models can be compared using the function \texttt{BLRT()}.

Anecdotal experience as a statistical co-author on LCA studies suggests that both these tests tend to be liberal,
returning significant p-values for models with more classes even when other indicators suggest that those models are overfitted (e.g., near-zero class counts).
Other authors have reported similar experiences (Sinha, Calfee, \& Delucchi, 2021), which matches the differences between LRT and ICs from information theory as LRT tests and compare the models based on in sample predictive accuracy, and ICs evaluates and compare the models out-of-sample predictive accuracy which helps to control for overfitting (McElreath, 2020).
This suggests that, like existing ICs,
LR tests might not adequately balance fit and complexity in LCA.
Better solutions for class enumeration are thus an important target for future research.

A promising novel solution is the so-called \emph{lazy bootstrap},
which is much more efficient than the BLRT because it is based on sample statistics of model-implied data generated from the \(k-1\) and \(k\)-class models,
rather than on fit statistics of LCA models fit to those data (Kollenburg, Mulder, \& Vermunt, 2018).
This simulation-based approach to inference is similar to the posterior predictive p-value in Bayesian statistics,
which describes how well each model is able to reproduce the observed correlation matrix (see McElreath, 2020).
One version of the lazy bootstrap is already implemented in the commercial program \emph{Latent GOLD} (Vermunt \& Magidson, 2021).
At the time of writing, we are developing a more general simulation-based likelihood ratio test (SBLRT) for class enumeration.
Future versions of \texttt{tidySEM} will likely incorporate this test.

@@

Classification diagnostics assess whether the classes identified by LCA are distinct and clearly separable.
These diagnostics are not model fit indices and should not be used for class enumeration,
because a model can fit the data well and still show poor latent class separation (Masyn, 2013).
However, classification diagnostics can be used to eliminate models:
If one analysis goal is to interpret properties of (members of) classes,
it makes sense to only consider models that classify well (Nylund-Gibson \& Choi, 2018).
Classification diagnostics should thus not be used for model selection, but they can help restrict the search space of acceptable models.

All classification diagnostics are derived from the posterior classification probabilities: an \(n \times k\) matrix called \(P\), with the probability that every individual \(n\) belongs to latent class \(k\).
When classification accuracy is high, and classes are distinct,
each individual's posterior class probabilities are high for one class, and low for the remaining classes.
The individual classification probabilities \(P\) are obtained by running \texttt{class\_prob(res,\ type\ =\ "individual")}.
Note that one column is added to the matrix, with each individual's most likely class membership \(M\).
\(M\) is obtained by assigning each individual to the class corresponding to their maximum posterior class probability.
Predicted class membership \(M\) is often of interest to researchers who wish to carry out follow up analyses.
Note, however, that follow-up analyses should take classification error into account.
The function \texttt{BCH()} does this automatically.

@@

To assess classification accuracy, we turn to the third diagnostic table:
The average posterior probabilities by most likely class membership, \(P_{p_{(M = m)}}\), \texttt{class\_prob(res,\ type\ =\ "avg.mostlikely")}:
for participants assigned to one particular class (rows) according to \(M\),
this table gives the mean probability of being assigned to any class (columns).
This table is obtained by calculating the column means of \(P\) for the
\(k\) subsets of observations with most likely class of \(M = 1 \ldots k\).
The diagonal represents the average predictive probability of a the subjects related to their most likely class, indicating the certainty of class assignment for the subjects. Some guidelines have being suggested for this diagonal, like for them to be higher than 0.7 (Masyn, 2013), but we suggest researchers to be cautions about these guidelines as they cannot be generalized and should not be use as strict cutoffs as evidence ``good'' or ``bad'' classification.

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{latent class analysis, mixture models, best practices, free open source software, tidySEM\newline\indent Word count: 12158}
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Best Practices in Latent Class Analysis using Free Open Source Software},
  pdfauthor={Caspar J. van Lissa1,2, Mauricio Garnier-Villarreal3, \& Daniel Anadria1},
  pdflang={en-EN},
  pdfkeywords={latent class analysis, mixture models, best practices, free open source software, tidySEM},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

LCA can also be used as a measurement model, similar to confirmatory factor analysis.
This is useful when class membership is measured imperfectly by several indicators:
LCA can partial out measurement error and restore most likely class membership.
For example, one study found discrepancies between self-reported employment status and official register data (Pavlopoulos \& Vermunt, 2015).
While it might be tempting to assume that register data is free of error, this assumption was shown to be false.
A mixture model was used to estimate employment status, correcting for measurement error, from both self-report and registry data.
The study investigated random and systematic measurement error in the indicators stemming from the two registers,
which revealed strengths and weaknesses of both types of assessment.

@@

The use of LCA for dimension reduction is common in longitudinal research,
where one developmental process is summarized using \emph{latent class growth analysis} (\emph{LCGA}),
and the resulting categories are then used as a moderator of another developmental process in a multi-group model.
For example, one study conducted a LCGA of adolescent empathy development on two dimensions of empathy identifying: high, average, and low empathy classes (Van Lissa et al., 2015).
The membership to these three groups was then used as a moderator in a latent growth analysis of adolescent- and parent-reported conflict.
The results showed that the high-empathy group evidenced greater adolescent-parent agreement about the incidence of conflict than other groups,
and the low empathy group had more conflict with parents than other groups.
Note that as both of these effects were non-linear,
modeling them in a single-group framework would introduce unnecessary bias.
LCGA greatly simplifies model specification and interpretation,
albeit at a cost of some loss of information, and potentially generalizability.
Appendix C is a tutorial on LCGA,
based on ongoing research on trajectories of depressive symptoms in military service members after deployment.

The need for dimension reduction can be especially pertinent when using ordinal indicators.
For example, one study measured
fifteen dichotomous health indicators among injured military personnel (MacGregor et al., 2021).
Combinatorics informs us that fifteen
dichotomous items have \(2^{15}\) or \(32,768\) unique symptom combinations.
However, LCA was able to reduce response patterns to five clinically meaningful latent classes supported by the data,
which makes intervention-planning significantly more manageable.

@@

\hypertarget{violations-of-normality}{%
\subsubsection{Violations of Normality}\label{violations-of-normality}}

Finally, LCA can be used to deal with data which violate certain assumptions.
For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) (Baughman, 2007; Magnus \& Garnier-Villarreal, 2022).
These models estimate one class to explain the ``true zeros'',
and another class with a different distribution to explain all other values, which can include zero.
Note that this LCA has different models for both classes,
and thus falls beyond the scope of this paper.
For instance, the aforementioned study on marine plastic noted that many prior studies had treated the distribution of particles as (multivariate) normal, despite clear deviations from normality (Alkema et al., 2022).
A mixture of two normal distributions with distinct variances described the observed distribution better.

@@

The best practices in estimation, as outlined in Table
\ref{tab:checkest-tab}, are rooted in existing recommendations
for estimating specific subtypes of LCA,
including latent
class growth analysis (Van De Schoot, Sijbrandij, Winter, Depaoli, \& Vermunt, 2017) and latent
class analysis with ordinal indicators (e.g., Nylund-Gibson \& Choi, 2018).
These guidelines were generalized to be relevant to all types of LCA,
and updated to current best practices, as explained below.

@@

\caption{\label{tab:checkest-tab}}
\centering
\begin{tabular}[t]{l|l}
\hline
\# & Item\\
\hline
1. & Examining Observed Data\\
\hline
2. & Handling Missing Data\\
\hline
3. & Alternative Model Specifications\\
\hline
4. & Software\\
\hline
5. & Algorithm\\
\hline
6. & Class Enumeration\\
\hline
7. & Model Fit Indices\\
\hline
8. & Classification Diagnostics\\
\hline
9. & Interpreting Class Solutions\\
\hline
10. & Label switching\\
\hline
\end{tabular}
\end{table}

@@

Examining observed data is essential for any analysis as it may reveal
patterns and violations of assumptions that had not been considered
prior to data collection.
The function \texttt{descriptives(data)} provides extensive descriptive statistics,
including the number of unique values,
variance of continuous and categorical variables,
their missingness, skew and kurtosis.
However, sample-level descriptive statistics are of limited value when the goal of a study is to identify subsamples using LCA.
Plots convey additional information; specifically, density plots for continuous variables, and bar charts
for categorical ones.
Density plots can also aid in the choice of the number of classes, as explained in the Visualization section.

Particular points of attention include measurement level of the indicators.
Recall that \texttt{tidySEM} can estimate mixture models with either numeric (\texttt{numeric} or \texttt{integer}) or \texttt{ordered} type indicators.
The column \texttt{type} should thus only contain these types.
Also check that the number of \texttt{unique} values match the intended measurement level.
Numeric variables with few unique values are typically better modeled as ordinal.
Our experience consulting on LCA methods
and moderating the \texttt{tidyLPA} Google group suggest that the
misapplication of mixture models to ordinal (e.g., Likert-type)
indicators is the most common source of user error. Whereas it has been
argued that some parametric methods are robust when scales with 7+
response categories are treated as continuous {[}e.g., Norman (2010); rhemtulla\_when\_2012{]},
it is very unlikely
that an ordinal variable can be modeled as a \emph{mixture} of multiple
normal distributions.
Consider what happens when estimating a 5-class solution on a single 5-point Likert scale indicator.
Each class-specific mean would describe a single response
category,
and a class-specific variance component would be empirically underidentified.
In sum, variables with few unique values might not be suitable for Gaussian mixture modeling.
LCA with ordinal indicators is more appropriate.

\hypertarget{data-preprocessing}{%
\subsubsection{Data Preprocessing}\label{data-preprocessing}}

LCA can accommodate variables of different measurement levels: continuous, binary, nominal, and ordinal.
Numeric variables should have an interval measurement level,
such that it is sensible to estimate theirmeans and (co)variances.
For the remaining three measurement levels,
some preprocessing is required.
The exact data types accepted by \texttt{tidySEM} at the time of writing are \texttt{numeric} and \texttt{ordered}.
To correctly specify ordered variables
such as Likert scales,
use the function \texttt{OpenMx::mxFactor()}.
Binary variables,
such as presence or absence of some symptom, are also treated as ordinal, with one category coded as zero and the other as one.
Nominal variables are converted to binary indicators of group membership via dummy coding, for example, using \texttt{mx\_dummies()}.

@@

A third consideration in preprocessing relates to the univariate distributions of indicators.
For example, the \texttt{plas\_depression} data shows extreme right skew in depressive symptoms:
most respondents report few symptoms,
and few respondents report many symptoms.
Similarly, the \texttt{alkema\_microplastics} data are extremely right-skewed: Very small plastic fragments abound in the oceans, and large pieces are rare.
These distributional characteristics must be taken into account either in preprocessing, or model specification.
In preprocessing, the data can be transformed to reduce skew.
After reducing skew, a simpler LCA model may fit the data well.
Alternatively,
one can specify a model that accommodates the distributional idiosyncrasies.
For example, allowing free variances across classes might result in one class with a near-zero mean and small variance to account for the fact that most respondents have few depressive symptoms,
and one class with an elevated mean and large variance to account for all respondents with non-zero symptoms.

@@

Previous work has emphasized the importance of examining the pattern of
missing data and reporting how missingness was handled
(Van De Schoot et al., 2017).
While we concur that investigating missingness is due diligence,
it is important to emphasize that missingness is adequately handled by default in \texttt{OpenMx}, which is the computational backend of \texttt{tidySEM}.
Its Full Information Maximum Likelihood (FIML) estimator makes use of all available
information without imputing missing values, by estimating the model based on all
available information for each subject (C. K. Enders, 2022; C. Enders \& Bandalos, 2001).
FIML is a best-practice solution for handling missing data;
at least on par with multiple imputation (Lee \& Shi, 2021).
Multiple imputation is less suitable to LCA for two reasons. First,
because LCA methods are often computationally intensive,
and conducting them on several imputed datasets multiplies the load.
Second, because many multiple imputation methods assume normality; an assumption likely to be violated when data come from a heterogeneous population, the imputation model could bias the results as it may account for the expected heterogeneity.

@@

The second consideration in model specification is the risk of overfitting (Hastie, Tibshirani, \& Friedman, 2009).
Overfitting occurs when a model has many parameters relative to the number of observations,
causing it to capture idiosyncratic noise in the sample and limiting its generalizability to new samples.
As explained before, the number of parameters in LCA scales with the number of estimated classes.
Consequently, LCA with many classes have a potentially very high number of parameters.
In Bayesian mixture modeling,
it was observed that if the estimated number of classes exceeded the true number, the number of observations assigned to these classes tended towards zero as sample size increased (Rousseau \& Mengersen, 2011).
Although it is not known whether frequentist LCA exhibits the same behavior,
low class counts are still an indicator of potential overfitting.
The section on Class Enumeration addresses ways to guard against overfitting.

@@

Due to the lack of absolute fit indices for LCA, the assessment of relative fit
relies on information criteria.
Information criteria (ICs) are computed from the \(-2LL\),
which is the most basic fit measure, testing the overall hypothesis if the model properly represents the data.
In addition to capturing model fit, ICs add a penalty for model complexity thus
preferring more model fit and parsimony.
The lower the value of an information criterion, the better the overall fit of the
model.
The \emph{Akaike Information Criterion} (\emph{AIC}) is the original information criterion.
It is computed as \(-2LL+2p\), where \(p\) is the number of parameters.
The \emph{Bayesian Information Criterion} (\emph{BIC}) multiplies the penalty for complexity with a function of the sample size, \(p \ln(n)\),
such that complex models are penalized more in smaller samples.
Finally,
the so-called \emph{sample size adjusted BIC} (\emph{saBIC}) implements a different ad-hoc penalty for sample size,
\(\ln(\frac{n + 2}{24})\). Although simulation studies have showen that the saBIC performs well for class enumeration (Chen, Luo, Palardy, Glaman, \& McEnturff, 2017; Tein, Coxe, \& Cham, 2013),
a limitation is that its penalty is not based on theory. BIC also performs well and is theoretically substantiated.
As such, BIC may be the most appropriate information criterion to use for model comparison (Masyn, 2013; Nylund-Gibson \& Choi, 2018).
In \texttt{tidySEM}, all three ICs are available as a part of the output of \texttt{table\_fit()}.

ICs may occasionally contradict each other, so it is
important to identify a suitable strategy to reconcile them.
One option is to select a specific fit index, such as the BIC,
before seeing the results.
This choice can be preregistered to avoid concerns about cherry picking.
Another option is to always prefer the most parsimonious model that has best fit according to any of the available fit indices.
Alternatively, one may select several competing models based on information criteria,
and evaluate them based on the theoretical interpretation of the identified classes.
Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process (Akogul \& Erisoglu, 2016),
which is implemented in \texttt{tidyLPA} (Rosenberg, Beymer, Anderson, Van Lissa, \& Schmidt, 2018).
Finally, one can use a scree plot to determine the inflection point at which additional classes contribute little to further decreases of the ICs (Nylund-Gibson \& Choi, 2018).
The preferred number of classes can be found at the inflection point where the slope of the curve is clearly leveling off.
Adding this many classes results in substantial decreases of the ICs,
and adding additional classes may add too much complexity to the model while not lowering the IC value.

Given a specific information criterion, it is also possible to calculate so-called ``IC-weights'',
which indicate the relative support the data provides for each model in the set (Wagenmakers \& Farrell, 2004).
This is particularly useful for confirmatory LCA,
as it allows one to quantify the support for the theoretical model.
In \texttt{tidySEM}, IC-weights are calculated by calling \texttt{ic\_weights()} on the model fit table.
Note, however, that IC-weights will (strongly) favor the model with the lowest IC value;
the scree plot approach above may be more suitable to select the most parsimonious model with adequate fit.

@@

When two models are nested, the significance of the difference in fit between these models can be tested using a \emph{Likelihood Ratio} (\emph{LR}) \emph{test}.
The test statistic is computed as the -2 times the log of the ratio of likelihoods,
and its degrees of freedom are equal to the difference in the degrees of freedom of the two compared models.
Since a \(k-1\)-class model is nested in a \(k\)-class model,
it is possible to test the difference in model fit using a LR test.
One challenge is that the LR test statistic is not asymptotically \(\chi^2\) distributed for LCA.
The so-called \emph{Lo-Mendell-Rubin (LMR) adjusted LR test} overcomes this problem by applying an ad-hoc correction to the LR (Lo, Mendell, \& Rubin, 2001).
This test is implemented in \texttt{tidySEM}, and is a part of the default output of \texttt{table\_fit()}.

Another solution is to derive an empirical sampling distribution for the LR difference.
The so-called \emph{Bootstrapped Likelihood Ratio Test (BLRT)} does so by simulating data from the \(k-1\)-class model \(b\) times (e.g., 1000),
and fitting the \(k\)- and \(k-1\)-class models to each simulated dataset (Nylund et al., 2007).
The p-value of this test is the proportion of \(b\) in which the LR of the models fit to simulated data exceeds the LR observed in the real data.
Simulation studies suggest that the BLRT has greater statistical power than the LRT,
although both perform comparably well at detecting the true number of classes (Nylund et al., 2007; Tein et al., 2013).
A crucial limitation of the BLRT is that it is very computationally expensive.
With this in mind, \texttt{table\_fit()} does not provide the BLRT by default.
Specific models can be compared using the function \texttt{BLRT()}.

The first author's anecdotal experience as a statistical co-author on LCA studies suggests that both aforementioned tests tend to be liberal,
returning significant p-values for models with more classes even when other indicators suggest that those models are overfitted (e.g., near-zero class counts).
Other authors have reported similar experiences (Sinha, Calfee, \& Delucchi, 2021), which matches the differences between LRT and ICs from information theory as LRT tests and compare the models based on in sample predictive accuracy, and ICs evaluates and compare the models out-of-sample predictive accuracy which helps to control for overfitting (McElreath, 2020).
This suggests that, like existing ICs, LR tests might not adequately balance fit and complexity in LCA.
Better solutions for class enumeration are thus an important target for future research.

A promising novel solution is the so-called \emph{lazy bootstrap},
which is much more efficient than the BLRT because it is based on sample statistics of model-implied data generated from the \(k-1\) and \(k\)-class models,
rather than on fit statistics of LCA models fit to those data (Kollenburg, Mulder, \& Vermunt, 2018).
This simulation-based approach to inference is similar to the posterior predictive p-value in Bayesian statistics,
which describes how well each model is able to reproduce the observed correlation matrix (see McElreath, 2020).
A version of the lazy bootstrap is already implemented in the commercial program \emph{Latent GOLD} (Vermunt \& Magidson, 2021).
At the time of writing, we are developing a more general \emph{simulation-based likelihood ratio test} (\emph{SBLRT}) for class enumeration.
Future versions of \texttt{tidySEM} will likely incorporate this test.

@@

Classification diagnostics assess whether the classes identified by LCA are distinct and clearly separable.
These diagnostics are not model fit indices and should never be used for class enumeration.
This is because a model can fit the data well and still exhibit poor separation between the latent classes (Masyn, 2013).
Instead, classification diagnostics can be used to eliminate models:
when the goal is to interpret properties of classes or their members,
it makes sense to only consider models which are successful at separating latent classes (Nylund-Gibson \& Choi, 2018).
Classification diagnostics should thus not be used for model selection, but they can help restrict the search space of acceptable models.

All classification diagnostics are derived from the posterior classification probabilities: an \(n \times k\) matrix called \(P\), with the probability that every individual \(i\) belongs to latent class \(k\)
When classification accuracy is high, and classes are distinct,
each individual's posterior class probabilities are high for one class, and low for the remaining classes.
In \texttt{tidySEM}, individual classification probabilities \(P\) are obtained by running \texttt{class\_prob(res,\ type\ =\ "individual")}.
Note that one column is added to the matrix, with each individual's most likely class membership \(M\).
\(M\) is obtained by assigning each individual to the class corresponding to their maximum posterior class probability.
Predicted class membership \(M\) is often of interest to researchers who wish to carry out follow up analyses.
Note, however, that follow-up analyses should take classification error into account.
The function \texttt{BCH()} does this automatically.

@@

To assess classification accuracy, we turn to the third diagnostic table:
the average posterior probabilities by most likely class membership, \(P_{p_{(M = m)}}\)
which can be obtained using \texttt{class\_prob(res,\ type\ =\ "avg.mostlikely")}.
For participants assigned to one particular class (rows) according to \(M\),
this table gives the mean probability of being assigned to any class (columns).
This table is obtained by calculating the column means of \(P\) for the
\(k\) subsets of observations with most likely class of \(M = 1 \ldots k\).
The diagonal represents the average predictive probability of the subjects related
to their most likely class, indicating the certainty of class assignment.
Some guidelines have being suggested for this diagonal,
such as its values having to be higher than 0.7 (Masyn, 2013).
We, however, advise caution as these guidelines cannot be generalized to all situations,
and should not be used as strict cutoffs communication the classification quality.
